---
title: "Principal Component Analysis in R"
author: "Aedin Culhane"
date: "July 15, 2020"
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
---


```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
# Set up the environment
opts_chunk$set(cache.path='cache/', fig.path='fig/', cache=FALSE, tidy=FALSE, fig.keep='high', echo=TRUE, dpi=100, warnings=FALSE, message=FALSE, comment=NA, warning=FALSE, results='as.is', fig.width = 10, fig.height = 6) 
```

### Necessary R libraries

```{r libraries}
# Install, if necessary, and load necessary libraries and set up R session
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
library(magrittr)
library(ade4)
library(FactoMineR)
library(factoextra)
library(ggplot2)
```

# PCA in R
In R, there are several functions in many different packages that allow us to perform PCA. We will not review all of these, however will provide examples of the following;

* `svd()` (`stats`)  ** on centered data** 
* `prcomp()` (`stats`)
* `princomp()` (`stats`)  ** on cor matrix **
* `PCA()` (`FactoMineR`)
* `dudi.pca()` (`ade4`)

**Note, although `prcomp` sets `scale=FALSE` for consistency with S, in general scaling is advised. We will demonstrate first pca of unscaled and then scaled data. Scaling scaled the variables to have unit variance and is advised before  analysis takes place.** 

We will demonstrate some of these and explore these using exploR


# Equivalents across methods

Give an input matrix P and result res

|Function | loadings | scores | plot|
| :------------- |:-------------| :-----| :-----|
|`prcomp(P, center=TRUE, scale=TRUE`) | `res$rotation` |`res$x` | `biplot(res)`|
|`princomp(P, cor=TRUE)` | `res$loadings` | `res$scores` | `biplot(res)`|
|`PCA(P)` | `res$svd$V` | `res$ind$coord` | `plot(res)`|
|`dudi.pca(P, center=TRUE, scale=TRUE)` | `res$c1` | `res$li` | `scatter(res)`|


With `ade4::dudi.pca` and `prcomp` the default is `center = TRUE, scale = TRUE`. 

With `princomp`, `cor=FALSE` is the default.

#  Create a simple toy dataset
Create a cloud of points; two vectors, x,y of length 100. 

```{r}
 set.seed(2)             #sets the seed for random number generation.
 x <- 1:100              #creates a vector x with numbers from 1 to 100
 ex <- rnorm(100, 0, 30) #100 normally distributed random numbers, mean=0, sd=30
 ey <- rnorm(100, 0, 30) # 100 normally distributed random numbers, mean=0, sd=30
 y <- 30 + 2 * x         #sets y to be a vector that is a linear function of x
 x_obs <- x + ex         #adds "noise" to x
 y_obs <- y + ey         #adds "noise" to y
 
# Bind both vectors in a matrix of toy data called P
P <- data.frame(x_obs=x_obs,y_obs=y_obs) #places points in matrix
summary(P)
```

Basic histogram of the two distributions:

```{r}
 par(mfrow=c(1,2))
 hist(P$x_obs)
 hist(P$y_obs)
```
 

Plot x,y. Place an "orange" point at the center (`mean(x)`, `mean(y)`) on plot

```{r}                                                                          
plot(P,asp=1,col=1) #plot points
points(x=mean(x_obs),y=mean(y_obs),col="orange", pch=19) #show center
```
 
 

# Computing via svd of **unscaled**, centered, covariance matrix

PCA can be computed as a singular value decomposition of a column centered matrix. Therefore we first processs the matrix. In this example, we don't scale. This is not advised.

```{r}
#center matrix
M <- cbind(x_obs-mean(x_obs),y_obs-mean(y_obs))
Mx<- scale(P, center=TRUE,scale=FALSE)
```

M equal to Mx, ignore col names
```{r}
all.equal(M, Mx, check.attributes=FALSE)  
```


# Eigenvector, Eigenvalues of the centered, covariance  matrix
The eigenvectors of the covariance matrix provide the principal axes, and the eigenvalues quantify the fraction of variance explained in each component. 

creates covariance matrix
```{r}
MCov <- cov(M) 
```

compute eigen values and vectors
```{r}
eigenvalues <- eigen(MCov)$values       
eigenvalues
```

(This is the same as `prcomp` PCA of the unscaled data)
```{r}
prcomp(P)$sdev^2 
```

and similar to `princomp`

```{r}
princomp(P)$sdev^2   
```


```{r}
eigenVectors <- eigen(MCov)$vectors     
eigenVectors
```

which is equivalent to 

```{r}
prcomp(P)$rotation
```

The right singular vectors are the eigenvectors of M^t^M. Next I plot the principal axes (yellow):
 
```{r}
plot(P,asp=1,col=1) #plot points
points(x=mean(x_obs),y=mean(y_obs),col="orange", pch=19) #show center
lines(x_obs,eigenVectors[2,1]/eigenVectors[1,1]*M[x]+mean(y_obs),col=8)
```

This shows the first principal axis. Note that it passes through the mean as expected. The ratio of the eigenvectors gives the slope of the axis. 

Next plot the second principal axis, orthogonal to the first

```{r}
plot(P,asp=1,col=1) #plot points
points(x=mean(x_obs),y=mean(y_obs),col="orange", pch=19) #show center
lines(x_obs,eigenVectors[2,1]/eigenVectors[1,1]*M[x]+mean(y_obs),col=8)
lines(x_obs,eigenVectors[2,2]/eigenVectors[1,2]*M[x]+mean(y_obs),col=8)
```
shows the second principal axis, which is orthogonal to the first (recall that the matrix V^t^ in the singular value decomposition is orthogonal). This can be checked by noting that the second principal axis is also, as the product of orthogonal slopes is -1. 

## PCA (scaled data) using Singular value decomposition  (svd)

Given a matrix X of dimension n × p, SVD decomposes it to:

$$M = U D V^{t}$$

U,V define the left and right singular values 

U and V are square orthogonal:

$$UU^{t} = I_{p}$$
$$VV^{t} = I_{n}$$

The output is 

d	- a vector containing the singular values 
u	- the left singular vectors 
v	- the right singular vectors of x

Relating to `prcomp`;

The columns u from the SVD correspond to the principal components x in the PCA. 

Furthermore, the matrix v from the SVD is equivalent to the rotation matrix returned by prcomp.

Now repeat the code above but **scale** and **center** the data with `scale(P, center=TRUE, scale=TRUE)`.

```{r}
any(M == scale(P))  #FALSE
all(scale(P, center=TRUE, scale=TRUE)== scale(P)) #TRUE
```

### `scale(P, center=TRUE, scale=TRUE)`
```{r}
p0<-svd(scale(P))

p0$d          #the singular values
p0$v          #the right singular vectors
```

The eigenvalues are
```{r}
diag(p0$d)
```

Which is 

```{r}
diag(t(p0$u) %*% M %*% p0$v)
```

Eigenvalues from svd on the scaled data. The diagonal elements of d from the SVD are proportional to the standard deviations (`sdev`) returned by PCA. 

The elements of d are formed by taking the sum of the squares of the principal components but not dividing by the sample size.

Therefore we can devide by the sample size, which is either the ncol or nrow of the matrix -1. 

```{r}
p0$d^2/(nrow(p0$u) - 1)
eigenValues= p0$d^2/(nrow(p0$u) - 1)
eigenValues
```

Summary of eigs
```{r}
eigSummary<-function(svdRes,sf=4){
  if(is(svdRes,"prcomp")){
    eigenvalue=svdRes$sdev^2
  }else{
  #d=signif(svdRes$d,sf)
  eigenvalue= svdRes$d^2/(nrow(svdRes$u) - 1)
}
  data.frame(cbind(
  eigenvalues=signif(eigenvalue,sf),
  sd = signif(sqrt(eigenvalue),sf),
  variance.percent = paste0(signif((eigenvalue/sum(eigenvalue)),2)*100,"%"),
  cumulative.variance.percent = paste0(cumsum(signif((eigenvalue/sum(eigenvalue)),2))*100,"%")))
}


eigSum.svd <-eigSummary(p0)
eigSum.svd 
```


## Demonstrating X = U D V' 
SVD returns a list with components

d	= a vector containing the singular values of x, of length min(n, p), in descending order

u	= a matrix whose columns contain the left singular vectors of x, present if nu > 0. Dimension c(n, nu).

v	= a matrix whose columns contain the right singular vectors of x, present if nv > 0. Dimension c(p, nv).

The product of these matrix X = U D V'  recovers the original matrix;
```{r}
(p0$u %*% diag(p0$d) %*% t(p0$v))%>% head
```
which equals the scaled matrix that was decomposed using svd

```{r}
scale(P)%>% head
```


## Comparing to prcomp

First `stats::prcomp`.  The eigenvectors are stored in `$rotation`. Note these are the same as `svd$v` on scale data 
```{r}
p1<- prcomp(P, scale = TRUE)
p1$rotation
```

```{r}
 (p1$rotation== p0$v)
```

eigenvalues - `$sdev`
eigenvector  - `$rotation`

```{r}
names(p1)
```

```{r}
summary(p1)
```

To calculated eigenvalues information manually here is the code;
```{r}
sf=4
eigs= p1$sdev^2
eigSum.pca= cbind(
  eigenvalue=eigs,
  sd = sqrt(eigs),
  variance.percent = eigs/sum(eigs),
  cumulative.variance.percent = cumsum(eigs)/sum(eigs))

eigSum.pca
```

```{r}
identical(eigSum.svd,eigSum.pca)
```


If we had more components, we could generate a scree plot. Its not very useful with 2 components, but here is the code

Caculate the Proportion of Variance explained by each component (eig sum Proportion above)
```{r}
ProportionVariance = p0$d^2 /sum(p0$d^2 )
ProportionVariance
```

```{r}
plot(ProportionVariance, xlim = c(0, 5), type = "b", pch = 16, xlab = "principal components", 
    ylab = "variance explained")
```

# `princomp`
`princomp` was written for compatiblity with S-PLUS however it is not recommended. Its is better to use prcomp or svd. That is because by default `princomp` performs a decompostion of the covariance not correlation matrix. `princomp` can call `eigen` on the correlation or covariance matrix. Its default calculation uses divisor N for the covariance matrix.

```{r}
p2<-stats::princomp(P)
p2$sd^2
```


sqrt of eigenvalues
```{r}
p2$sdev
```


eigenvectors
```{r}
p2$loadings
```

```{r}
head(p2$scores,2)
```

Set `cor = TRUE` in your call to princomp in order to perform PCA on the correlation matrix (instead of the covariance matrix)

```{r}
p2b<-princomp(P, cor = TRUE)
p2b$sdev^2
```

```{r}
p2b$loadings
```

For more info on `prcomp` v `princomp` see
http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/


## `FactoMineR`

`FactoMineR::PCA` calls svd to compute the PCA


```{r}
p3<-FactoMineR::PCA(P)
```

The eigenvalues, same as `eigSum` and `eigSum.svd` above
```{r}
t(p3$eig)
```

# correlations between variables and PCs
```{r}
p3$var$coord  
```


## `ade4::dudi.pca`

First `ade4::dudi.pca` scales the data and stores the scaled data in `$tab`. In PCA this will be almost equivalent to scale. However there is a minor difference (see https://pbil.univ-lyon1.fr/R/pdf/course2.pdf).  `ade4` uses the duality diagram framework for computing pca and other matrix factorizations (so it provides `lw` and `cw` which are the row and columns weights, respectively). See Cruz and Holmes 2011 for a wonderful tutorial on the duality diagram framework https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3265363/



```{r}
p4<-ade4::dudi.pca(P, scannf = FALSE, nf=2)  # save 2 axis by default,
head(p4$tab)  # centered/scaled data. 
head(scale(P))
```

The values used for centering are stored in `cent`, it is equal to the `colMeans`. `norm` provides the sd of the columns
```{r}
p4$cent == colMeans(P)
sd.n <- function(x) sqrt(var(x) * (length(x) - 1)/length(x))
identical(p4$norm,apply(P, 2, sd.n))
```
The summary printout is equivalent to P3 (`p3$eig`) above. 
The eigenvales are stored in `p4$eig`.

```{r}
summary(p4)
p4$eig
p4$c1
p4$co
```

The cumulative % of variance explained by each component:
```{r}
(k <- 100 * p4$eig/sum(p4$eig))
cumsum(k)
```

`nf` is an integer giving the number of axes kept. `nf` will always be smaller than the smaller of: the number of rows or number of columns of the matrix; minus 1.
```{r}
p4$nf
```

`c1` gives the variables’ coordinates, normed to 1. It is also called the
coefficients of the combination or the loadings of variables.  

Equally, the output matrix `l1` gives the individuals’ coordinates, normed to 1. It is also called the
loadings of individuals.
```{r}
p4$c1
sum(p4$cw * p4$c1$CS1^2)
```

`co` gives the variables’ coordinates, normed to the square root of the eigenvalues.
```{r}
p4$co
sum(p4$cw * p4$co$Comp1^2)
```


The link between `c1` and `co` is defined by:
```{r}
p4$c1$CS1 * sqrt(p4$eig[1])
```


# Z-score normalization

The default version of scale is a z-score standarization

```{r}
zscore<- function(x) (x-mean(x))/sd(x)
PZ<-apply(P,2, zscore)
all(PZ== scale(P))
data.frame(mean=round(colMeans(PZ)), sd= apply(PZ,2, sd))
```

Then by extension, a P0, svd of a scaled matrix is the same as svd of a z-score matrix

```{r}
pz<-svd(PZ)
all(pz$v == p0$v)
all(pzEigs<-pz$d^2/(nrow(pz$u) - 1) ==eigs)
```


## Comparision of results of different methods

There is also a nice package called `factoextra`. This works all of the above classes

```{r}
library(factoextra)

res<- list(p0,p1,p2,p2b,p3,p4) 
names(res) = c('svd_scaledData','prcomp', 'princomp','princomp_cov', 'FactoMineR', 'ade4')

e<-sapply(res[-1],get_eig)

# get_eig doesn't work on svd
svd.e<- eigSum.svd[c(1,3,4)]

colnames(svd.e)<- names(e[[1]])


e<- c(list(svd=svd.e),e)

e
```


# Visualization and Exploration of results

The github package [`explor`](https://github.com/juba/explor) is useful for exploring data. It includes plotting functions for many packages including `ade4`, `FactoMineR` and `base` R functions `prcomp` and `princomp`;

For now on, it is usable the following types of analyses :

Analysis | Function  | Package | Notes
------------- | ------------- | ---------- | --------
Principal Component Analysis  | PCA  | [FactoMineR](http://factominer.free.fr/) | -
Correspondance Analysis  | CA  | [FactoMineR](http://factominer.free.fr/) | -
Multiple Correspondence Analysis  | MCA  | [FactoMineR](http://factominer.free.fr/) | -
Principal Component Analysis  | dudi.pca  | [ade4](https://cran.r-project.org/package=ade4) | Qualitative supplementary variables are ignored
Correspondance Analysis  | dudi.coa  | [ade4](https://cran.r-project.org/package=ade4)  | -
Multiple Correspondence Analysis  | dudi.acm  | [ade4](https://cran.r-project.org/package=ade4) | Quantitative supplementary variables are ignored
Specific Multiple Correspondance Analysis | speMCA | [GDAtools](https://cran.r-project.org/package=GDAtools) | Supplementary variables are not supported
Multiple Correspondance Analysis | mca | [MASS](https://cran.r-project.org/package=MASS) | Quantitative supplementary variables are not supported
Principal Component Analysis  | princomp  | stats | Supplementary variables are ignored
Principal Component Analysis  | prcomp  | stats | Supplementary variables are ignored




```{r}
#if(!"explor" %in% rownames(installed.packages()))    #devtools::install_github("juba/explor")

#if(!"scatterD3" %in% rownames(installed.packages())) 
#devtools::install_github("juba/scatterD3")

```

```{r, eval=FALSE}
require(explor)
explor::explor(p4)
```


```{r, eval=FALSE}
data(children)
res.ca <- CA(children, row.sup = 15:18, col.sup = 6:8)
explor(res.ca)
```

## `factoextra`
Plotting using `factoextra`

```{r}
library(factoextra)
```

```{r}
fviz_eig(p1)
```

```{r}
fviz_pca_var(p1,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )

```

```{r}
fviz_pca_biplot(p1, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )

```



# Drawing Ellispe

Example using iris dataset

```{r}
data(iris)
ir.pca<-prcomp(log(iris[,1:4]), center=TRUE, scale=TRUE)
```

Easiest approach: 
```{r}
library(ggplot2)
library(ggfortify)
ggplot2::autoplot(ir.pca, data=iris, colour="Species", frame=TRUE, frame.type="t") 
```

```{r}
library(ggplot2)
ggplot(ir.pca,aes(PC1, PC2))+ 
  geom_point() + 
  stat_density_2d(aes(alpha=..level.., fill=iris$Species), bins=4, geom="polygon")
```



`stat_ellipse()` and `stat_density_2d()` have a lot of options. See manual pages

multivariate normal distribution.
```{}
stat_ellipse(type = "norm", linetype = 2)  
```

Euclid, is a circle with radius equal to the `level` parameter
```{}
stat_ellipse(type = "euclid", level = 3) 
```

multivariate t-distribution
```{}
stat_ellipse(type = "t") 
```




