---
title: 'PCA example: Single cell count data'
author: "Lauren Hsu"
date: "7/16/2020"
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{PCA example: Single cell count data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this example, we will 


## PCA

```{r}

```

### *Interactive exercise 1*

Write 

```{r}
# your code here
```

## Visualizing PCA



```{r}

```



## Speeding up PCA

This dataset is small compared to many that we encounter in biology. 

```{r}

```

As we saw above, we can achieve the same results as a `prcomp` call by using the `scale` and `svd` function.

Similarly, 

### *Interactive exercise 2*

Perform PCA on 

```{r}
# time comparison prcomp, scale+svd, scale+irlba
```


## Brief introduction: correspondence analysis

PCA is the oldest dimension reduction method. Nonetheless, it is still in widespread use today.



There are numerous methods related to PCA; one such method is **correspondence analysis**, which is similar to . 

Correspondence analysis is implemented in various other R packages, including:

- `ade4`::`dudi.coa()`
- `made4`::`ord(type = 'coa')`
- `vegan`::`cca()` *(not to be confused with canonical correlation analysis, also abbreviated CCA)*
- `FactoMineR`::`CA()`
- `MASS`::`corresp()`

We recently wrote a version
